Perturbative PDF of the total magnetization of the 4D Ising model

### 2020-04-28 paper/paper.tex
Proofread

\author[a]{Andrea Allais}
\affiliation[a]{Independent scholar}
\emailAdd{allais.andrea@gmail.com}

\abstract{We compute, at one loop in perturbation theory, the probability
density function of the total magnetization $M$ of the Ising model, on both the
4-torus and the 4-sphere. We develop a single perturbative expansion that is
valid both in the symmetric phase, as long as the correlation length $\xi$ is
large in lattice units, and in the broken symmetry phase, as long as the
correlation length is large compared to the system size $L$. We find that, at
the critical point, for very large system size in lattice units, the PDF
approaches $p(M)\sim \exp(-f(L) M^4)$, and hence the critical value of the
Binder cumulant of the total magnetization is $U = 1 -
\frac{4\,\Gamma(5/4)^2}{3\,\Gamma(3/4)^2}$. We validate our results by
comparison with Monte Carlo simulation.}
\begin{document}

@@

When the system size $L$ is large, the probability distribution of the total
magnetization $M$ of the Ising model changes qualitatively between the two
phases of the model. In the symmetric phase, \textit{i.e.} when the correlation
length $\xi$ is large in lattice units, but small compared to the system size,
the distribution is well approximated by a zero-mean normal.  In the broken
symmetry phase, still for $1 \ll \xi \ll L$, the distribution is well
approximated by a mixture of two normals centered at non-zero values $\pm
M_0$.

Near the critical point $1 \ll L \ll \xi$, in two and three dimensions, the
probability distribution of $M$ is a non-trivial function of $M$.  It is not
immediately clear if the same is also true in four dimensions. Since the field
theory that describes the Ising critical point in four dimensions becomes
weakly coupled at low energy, one may expect the probability distribution of
$M$ to remain a zero-centered normal even at the critical point. This is in
fact the case for the total magnetization of a subsystem of intermediate size:
large in lattice units, but small compared to $L$ \cite{Binder1981}. Here we
show that the same is not true for the total magnetization of the entire
system. We compute the logarithm of the PDF of $M$ at one-loop in $4$
dimensions, and show that, for very large system size, the critical probability
distribution is not normal, but rather of the form $-\log p(M)\sim f(L) M^4$.
We find that corrections to this form vanish very slowly with increasing system
size, like $1/\log L$.

@@

Here we take the manifold $\mathcal{M}$ to be a 4-torus of radius $R$,
\textit{i.e.} $x_i\sim x_i + 2\pi R$. In this field theory context, we define
the PDF of the average magnetization $m = M / V$ as:
\begin{equation}
    \label{eq:pdf_definition}
    p(m) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
    \delta\left(m - \frac{1}{V}\int\dd^4 x\ \phi(x)\right)
    e^{-S[\phi]}\,,
\end{equation}
and we evaluate it at one loop in perturbation theory, obtaining:
\begin{equation}
\label{eq:log_pdf}
\begin{split}
    -\log p(m) = \mathrm{const.} +  V\Bigg(
  &\frac{1}{2} \frac{m^2}{R^2} \left(
    r R^2 + \frac{g}{6} \left(f_1\left(r R^2\right) - 1\right)
    + O\left(g^2\right)\right) + \\
  & \frac{2\pi^2}{9}gm^4\left(
    1 - \frac{g}{2}f_2\left(r R^2\right)
    + O\left(g^2\right)\right) + \\
  & \frac{8\pi^4}{81} g^3 m^6 R^2\left(
    f_3\left(r R^2\right)
    + O\left(g\right)\right) + O\left(g^4m^8\right) \Bigg)\,.
\end{split}
\end{equation}
Here the coefficients $r$ and $g$ are the renormalized counterparts to $r_0$
and $u_0$; the functions $f_1$, $f_2$ and $f_3$ are plotted in
fig.~\ref{fig:f_functions}, and an explicit expression (\ref{eq:f_functions})
is given below.

@@

This is our central result, and we now discuss it at length. We first
illustrate its domain of validity and its dependence on the renormalized
couplings $r$, $g$ on a torus of fixed radius $R$, then describe its dependence
on $R$ at fixed bare couplings $r_0$, $u_0$.

@@

At the critical point, the renormalized coupling $g$ follows the simplified
Callan-Symanzik equation
\begin{equation}
    \frac{1}{g(R_2)} - \frac{1}{g(R_1)} = \log\frac{R_2}{R_1}\,,
\end{equation}
and hence, as system size grows, the renormalized coupling $g$ goes to zero as
$1 / \log(R)$.  In this regime, the quartic term in (\ref{eq:log_pdf})
dominates all the others. This is perhaps most evident if the PDF is expressed
in terms of the rescaled quantity $\bar{m} = g^{\frac{1}{4}} m$, whose variance
remains finite as $g\to0$. Thus we conclude that, at the critical point, for
sufficiently large system size:
\begin{equation}
    \label{eq:log_pdf_critical_point}
    -\log p(m) \sim \mathrm{const} + \frac{2\pi^2}{9} V gm^4\,.
\end{equation}

@@

Retaining only one-loop diagrams, and employing a heath kernel regulator we have:
\begin{equation}
\begin{split}
    -\log p(m) = \mathrm{const} + V \Bigg( 
    &\frac{1}{2} m^2\left( r + u \left(\delta r + \frac{1}{2} I_1\right) + O(u^2)\right) + \\
    &\frac{u}{24} m^4 \left(1 + u \left(\delta u - \frac{3}{2} I_2\right) + O(u^2)\right) + \\
    &\frac{u^3}{48}m^6\left(I_3 + O(u)\right) + O(u^4 m^8) \Bigg)\,,
\end{split}
\end{equation}
where
\begin{equation}
    I_k = \frac{1}{V} \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    \frac{e^{-s(n^2 R^{-2} + r)}}{\left(n^2 R^{-2} + r\right)^k} = 
    \frac{1}{(k - 1)!}\left(-\frac{\partial}{\partial r} - s\right)^{k-1} I_1
\end{equation}

>>>>

\author[a]{Andrea Allais}
\affiliation[a]{Independent scholar}
\emailAdd{a\_allais@alum.mit.edu}

\abstract{We compute, at one loop in perturbation theory, the probability
density function of the total magnetization $M$ of the Ising model on the
4-torus and the 4-sphere. We develop a single perturbative expansion that is
valid both in the symmetric phase, as long as the correlation length $\xi$ is
large in lattice units, and in the broken symmetry phase, as long as the
correlation length is large both in lattice units and compared to the system
size $L$. We find that, at the critical point, for large system size in lattice
units, the PDF approaches $p(M)\sim \exp(-f(L) M^4)$. Consequently, the
critical value of the Binder cumulant of the total magnetization is $U = 1 -
\frac{4\,\Gamma(5/4)^2}{3\,\Gamma(3/4)^2}$. We validate our results by
comparison with Monte Carlo simulation.}
\begin{document}

@@

When the system size $L$ is large, the probability distribution of the total
magnetization $M$ of the Ising model changes qualitatively between the two
phases of the model. In the symmetric phase, if the correlation
length $\xi$ is large in lattice units, but small compared to the system size,
the distribution is well approximated by a zero-mean normal.  In the broken
symmetry phase, still for $1 \ll \xi \ll L$, the distribution is well
approximated by a mixture of two normals centered at non-zero values $\pm
M_0$.

In the opposite regime, $1 \ll L \ll \xi$, in two and three dimensions, the
probability distribution of $M$ is a non-trivial function of $M$.  It is not
immediately clear if the same is also true in four dimensions. Since the field
theory that describes the Ising critical point in four dimensions becomes
weakly coupled at low energy, one may expect the probability distribution of
$M$ to remain a zero-centered normal even in this regime. This is in fact the
case for the total magnetization of a subsystem of intermediate size: large in
lattice units, but small compared to $L$ \cite{Binder1981}. Here we show that
the same is not true for the total magnetization of the entire system. We
compute the logarithm of the PDF of $M$ at one-loop in $4$ dimensions, and show
that, for $1 \ll L \ll \xi$, the distribution is not normal, but rather of the
form $-\log p(M)\sim f(L) M^4$.  We find that corrections to this form vanish
very slowly with increasing system size, like $1/\log L$.

@@

Here we take the manifold $\mathcal{M}$ to be a 4-torus of radius $R$,
\textit{i.e.} $x_i\sim x_i + 2\pi R$. In this field theory context, we define
the PDF of the average magnetization $m = M / V$ as:
\begin{equation}
    \label{eq:pdf_definition}
    p(m) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
    \delta\left(m - \frac{1}{V}\int_{\mathcal{M}} \phi\right)
    e^{-S[\phi]}\,,
\end{equation}
and we evaluate it at one loop in perturbation theory, obtaining:
\begin{equation}
\label{eq:log_pdf}
\begin{split}
    -\log p(m) = \mathrm{const.} +  V\Bigg(
  &\frac{1}{2} \frac{m^2}{R^2} \left(
    r R^2 + \frac{g}{6} \left(f_1\left(r R^2\right) - 1\right)
    + O\left(g^2\right)\right) + \\
  & \frac{2\pi^2}{9}gm^4\left(
    1 - \frac{g}{2}f_2\left(r R^2\right)
    + O\left(g^2\right)\right) + \\
  & \frac{8\pi^4}{81} g^3 m^6 R^2\left(
    f_3\left(r R^2\right)
    + O\left(g\right)\right) + O\left(g^4m^8\right) \Bigg)\,.
\end{split}
\end{equation}
Here the coefficients $r$ and $g$ are the renormalized counterparts to $r_0$
and $u_0$; the functions $f_1$, $f_2$ and $f_3$ are plotted in
fig.~\ref{fig:f_functions}, and an explicit expression (\ref{eq:f_functions})
is given below.

@@

%This is our central result, and we now discuss it at length. We first
%illustrate its domain of validity and its dependence on the renormalized
%couplings $r$, $g$ on a torus of fixed radius $R$, then describe its dependence
%on $R$ at fixed bare couplings $r_0$, $u_0$.

@@

At the critical point, the renormalized coupling $g$ follows the simpler
Callan-Symanzik equation:
\begin{equation}
    \frac{1}{g(R_2)} - \frac{1}{g(R_1)} = \log\frac{R_2}{R_1}\,,
\end{equation}
and hence, as system size grows, the renormalized coupling $g$ goes to zero as
$1 / \log(R)$.  In this regime, the quartic term in (\ref{eq:log_pdf})
dominates all the others. This is perhaps most evident if the PDF is expressed
in terms of the rescaled quantity $\bar{m} = g^{\frac{1}{4}} m$, whose variance
remains finite as $g\to0$. Thus we conclude that, at the critical point, for
sufficiently large system size:
\begin{equation}
    \label{eq:log_pdf_critical_point}
    -\log p(m) \sim \mathrm{const} + \frac{2\pi^2}{9} V gm^4\,.
\end{equation}

@@

Retaining only one-loop diagrams, and employing a heat kernel regulator we have:
\begin{equation}
\begin{split}
    -\log p(m) = \mathrm{const} + V \Bigg( 
    &\frac{1}{2} m^2\left( r + u \left(\delta r + \frac{1}{2} I_1\right) + O(u^2)\right) + \\
    &\frac{u}{24} m^4 \left(1 + u \left(\delta u - \frac{3}{2} I_2\right) + O(u^2)\right) + \\
    &\frac{u^3}{48}m^6\left(I_3 + O(u)\right) + O(u^4 m^8) \Bigg)\,,
\end{split}
\end{equation}
where
\begin{equation}
    I_k = \frac{1}{V} \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    \frac{e^{-s(n^2 R^{-2} + r)}}{\left(n^2 R^{-2} + r\right)^k} = 
    \frac{1}{(k - 1)!}\left(-\frac{\partial}{\partial r} - s\right)^{k-1} I_1
\end{equation}
